Tree size, microhabitat, and hydraulic traits (interactively) shape drought response in a temperate broadleaf forest

### 2019-08-03 manuscript/climate_sensitivity_draft.Rmd
refined table/ table heading
re-wrote hypothesis text
Following addition of Table 1, it was possible to condense text to one paragraph (last paragraph of intro).

@@

@@

*Third, are responses similar or variable across individual drought years?*
Droughts are rarely explicitly defined in ecological studies [@slette_how_2019], yet no two droughts are the same. This study addressed this by analyzing trees' resistance to drought within and across three defined drought periods.

*Hypotheses*

**Table 1. Summary of hypotheses, predictions and results** We count predictions as fully supported (or rejected) when the direction of response matches (or contradicts) the prediction and model dAIC≥2 relative to the approriate null model or to any alternative multivariate model with within 2 dAIC. Parentheses indicate that predictions were partially supported (or rejected)-- i.e., with direction of response matching (or contradicting) the prediction but model dAIC<2 relative to the approriate null model, when response partially matched predicted order for categorical variable (canopy position), or when an alternative multivariate model within 2 dAIC had a contradictory response.  (Table 1.xlsx)

(The content below needs to be re-written / mostly removed to go with Table 1.)

**1. Why do larger trees suffer greater growth declines during drought?**

1.0. Consistent with most forests globally [@bennett_larger_2015], larger-diameter trees have lower drought resistance. (Note that Bennett et al. 2015 identified only one study on tree growth responses to drought in the Eastern US temperate deciduous biome. We know little about how tree size shapes drought response in this biome.)

  * P1.0-Drought resistance decreases with DBH at time of drought. (supported most of the time)

1.1. Drought resistance decreases with tree height.

  * P1.1- Drought resistance decreases with height at time of drought, and height is a better predictor than DBH. (supported most of the time)

1.2. Canopy trees have lower drought resistance because they are exposed to higher solar radiation, greater wind speeds, and lower humidity. Alternatively, the generally supressed status of subcanopy trees may be insufficient to override the benefits of their buffered environment during drought. 
  * P1.2a- Canopy position is a better predictor of resistance than height, with trees currently in a dominant canopy position having lowest resistance. (rejected)
  * P1.2b- After accounting for height, trees currently in a dominant canopy position have lowest resistance. (partially supported)

1.3. During drought, larger trees are at a relatively smaller disadvantage in drier microenvironments where rooting volume/depth is more important for reliable water access.

  * P1.3- There is a negative interactive effect between tree height and topographic water index. (rejected- [issue #45](https://github.com/SCBI-ForestGEO/McGregor_climate-sensitivity-variation/issues/45))

**2. How do species' traits--alone and in interaction with tree size--influence drought response?**

2.1. Commonly measured traits including wood density (REFS), leaf mass per area [@abrams_adaptations_1990] [@guerfel_impacts_2009], and ring porosity (Elliot et al. 2015, Friedrichs et al. 2009) have been linked to drought responses and likely correlated with drought resistance in this forest. However, we hypothesize that leaf hydraulic traits such as leaf area shrinkage upon dessication and turgor loss point, which are emerging as potentially more informative traits but whose effect on drought resistance has never been tested (CONFIRM), will prove better predictors.
    * P2.1 - wood density correlates positively drought resistant
    * P2.2 - leaf mass area correlates positively to drought resistance 
    * P2.3 - diffuse porous species are more sensitive than ring porous  
    * P2.4 -leaf area shrinkage upon dessication (PLA) is negatively correlated with drought resistance (supported as one of the best predictors)
    * P2.5 - TLP correlates negatively with drought resistance  (supported as one of the best predictors)

2.2. Traits vary with microenvironment, with more drought-resistant traits associated to taller trees (and drier topographic positions) ([issue #43](https://github.com/SCBI-ForestGEO/McGregor_climate-sensitivity-variation/issues/43))

  * P2.2a- traits (all 5 above) correlate with tree height 
  * (P2.2b- traits (all 5 above) correlate with TWI)

2.3. Size-dependent drought resistance is driven by--or alternatively, buffered by--functional traits (i.e., larger trees have more drought vulnerable traits).

  * P2.3 Inclusion of traits in the statistical model does not eliminate or significantly reduce effect of tree size (or TWI). That is, the effects of height and and microhabitat will not be negated in the best model including size, microhabitat, and traits.  (supported)

**3. Did responses vary across droughts?**
3.1 Overall community response is proportional to the severity of each drought.  (Do you really want to add this hypothesis? How do you define severity? min PDSI? duration? it may be tricky to find the right metric, but I agree that its interesting to compare across droughts)

3.2. Predictor variables varied across droughts.

  * P3.2 - There will be overall consistnecy in terms of the directions of responses to predictor variables; however, the best set of predictor variables will vary across droughts.

>>>>

Canopy trees have lower drought resistance because they are exposed to higher solar radiation, greater wind speeds, and lower humidity. Alternatively, the generally supressed status of subcanopy trees may be insufficient to override the benefits of their buffered environment during drought.

@@

Commonly measured traits including wood density (REFS), leaf mass per area [@abrams_adaptations_1990] [@guerfel_impacts_2009], and ring porosity (Elliot et al. 2015, Friedrichs et al. 2009) have been linked to drought responses and likely correlated with drought resistance in this forest. However, we hypothesize that leaf hydraulic traits such as leaf area shrinkage upon dessication and turgor loss point, which are emerging as potentially more informative traits but whose effect on drought resistance has never been tested (CONFIRM), will prove better predictors.

@@

*Third, are responses similar or variable across individual drought years?*
Droughts are rarely explicitly defined in ecological studies [@slette_how_2019], yet no two droughts are the same.

We need to understand the the factors confirming drought vulnerability or resistance not only for extreme droughts with dramatic impacts on tree growth and mortality, which tend to dominate the literature ([@bennett_larger_2015], REFS), but also for more modest but frequent droughts--e.g., those with historical return intervals on the order of a decade.

Here, we combine tree-ring records covering three droughts (1964-66, 1977, 1999), species functional and hydraulic trait measurments, and forest census data from a 25.6-ha ForestGEO plot in Virginia (USA) to test a series of hypotheses and associated specific predictions (Table 1) designed to yield functional understanding of how tree size, microenrivonment, and species' traits collectively shape drought responses. First, we focus on the role of tree size and its interaction with microenvironement. We confirm that, consistent with most forests globally [@bennett_larger_2015], larger-diameter trees have lower drought resistance in this forest, which is in an ecoregion (Eastern US temperate deciduous) represented by only one study in Bennett et al (2015) (H1.0). We then test hypotheses designed to disentangle the relative importance of tree height (H1.1), crown exposure (H1.2), and root water access, which should be greater for larger trees indry but not perpetually wet microsites (H1.3). Second, we focus on the role of species' functional and hydraulic traits and their interaction with tree height. We hypothesize that drought resistance will follow predicted and observed patterns in relation to wood density, specific leaf area, and ring porosity (P2.1a-c), but that leaf hydraulic traits such as leaf area shrinkage upon dessication and turgor loss point will prove better predictors (P2.1d-e). We then test whether these traits correlate with tree height (P2.2), potentially driving the observed tendency for taller trees to suffer more during drought (P2.3). Finally, we focused on variability among droughts, asking how  community resistance varied across droughts (H3.1) and whether the factors confirming vulnerability or resistance varied across droughts (H3.2).

**Table 1. Summary of hypotheses, predictions and results** We count predictions as fully supported (or rejected) when the direction of response matches (or contradicts) the prediction and model dAIC≥2 relative to the approriate null model or to any alternative multivariate model with within 2 dAIC. Parentheses indicate that predictions were partially supported (or rejected)-- i.e., with direction of response matching (or contradicting) the prediction but model dAIC<2 relative to the approriate null model, when response partially matched predicted order for categorical variable (canopy position), or when an alternative multivariate model within 2 dAIC had a contradictory response. "(yes/no) or ("no/yes")" indicates tendancies in opposite directions in univariate tests versus the best full models, respectively. (Table 1.xlsx)
